Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-07-31 Manuscript/MPESubmission/Revision/response.tex

\item{\textcolor{blue}{\textbf{line 286:} Why was the true tree used as a starting tree? Earlier the authors argued that the true tree is almost never known, so using it here detracts from the generalizability of these results. If the only justification is 'to speed up the Bayesian estimation process' as stated later, then it seems like this may bias one of the main results, which is that Bayesian analyses outperformed max. likelihood, in which no starting tree was used. I understand the computational limitations, but the authors should present some analysis with complete and missing data in which no starting tree is given compared to analyses in which a starting tree is given to evaluate whether using a starting tree significantly affects the resulting trees, perhaps in an Appendix. Or at least include some perturbation to the starting tree to avoid overly influencing the resulting topology. It would seem to me that this is illustrated in figure 5, where-in with the Robinson-Foulds metric, the distance to the best tree in Bayesian analysis seems to have a lower bound at ~0.7, even under the worst cases of missing data. This may be the worst the analysis can do given that it started out with the true tree. }}

We dealt with this comment above (see Reviewer 1 major comment 7).

@@

\end{enumerate}

\section{Reviewer 2:}
Note that we split the second comment from reviewer 2 in three different comments (comments 2,3 and 4 below) to improve the readability of our response to the reviewers.

\begin{enumerate}
\item{\textcolor{blue}{\textbf{Compare the trees to the "true" tree:} I don't see how the "best" tree approach is helpful in this case. It adds an additional layer of complexity that can't be accounted for via simulation. The point of simulations is to compare empirical realities to the generating process, to see how much results diverge. I think the most helpful and straightforward thing to do is eliminate the "best" tree part of the MS, and compare the inferred trees directly to the real topology to quantify variance.}}

A similar point/suggestion was also raised by reviewer 1 (see Reviewer 1 General comment number 6 and specific comment number 25).
We do understand the concerns of this reviewer as well but, again, we do not agree that comparing the "missing-data" trees to the "best" trees adds a layer of complexity to our simulations.
In fact, we argue that comparing the "missing-data" trees to the "true" tree actually adds a layer of complexity to the analysis.
The problematic we are tackling in this manuscript is to understand "what is the effect of our three parameters on recovering topology ?"
To answer this question, however, we have to add a layer between the data (i.e. the "complete" or "missing-data" matrices) and the actual topology which is the tree inference.
We test the effect of this layer by using different tree inference methods.

@@

\item{\textcolor{blue}{\textbf{Specific variation of data per fossil:} More importantly, it doesn't seem like missing data varied among fossil taxa, unless I'm misreading. What we really want to know is what happens when some fossil taxa are fragmentary, as this is the most common empirical situation. Thus, there are morphological characters scored for some living taxa, some fossil taxa, and missing from others. It's unclear if this was simulated. It would be good if the authors could detail the case where, say, all living taxa had morphological data (0 Ml), the matrix was large (0 Mc), and Mf varied *among* fossil taxa, such that some fossils had 100\% coverage, and varied down to, say, 5\%. That's the crucial case here.}}

In this study, we applied a similar approach to what the reviewer 2 suggests (i.e. no missing data for living taxa, and maximal number of characters but various amount of missing data for the fossil taxa).
This results are present in figure 4, central row.
However, the amount of changes among the fossil taxa was not set to any specific amount of missing data per fossil (e.g. like the "artificial fossil" approach in Pattinson et al 2014, Syst. Biol.) but we rather treated the missing data in the "whole" fossil record.
Because our fossil taxa and their characters where randomly simulated, we didn't had any \textit{a priori} assumption on how to remove the data.
We therefore decided to randomly remove a certain amount of data among all the fossil.
For example, when removing 75\% of the data for the fossil taxa ($M_F$=75\%) different scenarios could happen, ranging from a "homogeneous" incompleteness data distribution (each specific fossil taxa had 75\% of missing data) to a "heterogeneous" incompleteness data distribution  (80\% of the specific fossil taxa had 5\% available data and 21\% of the specific fossil taxa had 100\% available data).
In empirical matrices, the incompleteness data distribution probably lies in between both models.

Also, a second argument for using this "simplistic" way to deal with the missing data in the fossil record follows our main recommendation from this study: the best way to improve the quality of Total Evidence trees is to have a maximum number of living species with available morphological data.
In fact lines 584 to 589 in the first version of the manuscript (line @@@ in the revised version), we argue that even if the effect of the $M_F$ parameter is important, there is unfortunately not much we can do about it since the completeness of this part of the matrix solely depends on exceptional palaeontological discoveries and excavation effort.
We feel that these parameters, however crucial for our understanding of macroevolution, are unfortunately not predicable and a recommendation such as "we need more fossil data" is less practical than our main recommendation: "we need more living species data".
